On Optimality of Finite Element Integration

### 2015-11-30 TOMS_Paper/assembly.tex
trying to write structured tensor operations

\begin{Def}[Quasi-optimality of a loop nest]
\label{def:mln-quasi-optimality}
Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in the innermost loop.
\end{Def}

@@

Achieving optimality in polynomial time is clearly not always possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily complex. Multilinearity, on the other hand, ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions. In the following sections, we build an approach that achieves: (i) optimality whenever the $\sigma_{ei}$  expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases. In particular, we need:
\begin{itemize}
\item the space of legal transformations impacting the operation count
\item an algorithm to select points in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space and we explain the conditions for optimality. We show that whenever an expression (i.e., its monomials) is ``too unstructured'', optimality may not be reached. We will consider multilinear loop nests of arity $n=2$ (i.e., bilinear forms), although a similar (simpler) reasoning straightforwardly applies to the case $n < 2$.

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, code motion and (if necessary) common sub-expression elimination. If the objective is reaching optimality and expressions are generally unstructured, transformations of this sort require solving large combinatorial problems, for instance to evaluate the impact of all possible factorization strategies. In our domain, however, expressions may be characterized by certain properties. These properties hold either in general -- as inherent in finite element integration -- or in forms making ``special use'' of most common differential and tensor operators. We exploit them to build an algorithm that results in optimal or quasi-optimal sharing elimination depending on the problem to which it is applied. In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss these properties, while in Section~\ref{sec:sec:se-algo} we present the algorithm. We assume the usual finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in the canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

\subsubsection{Common sub-expressions}
\label{sec:se-mln}
A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE). This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.

We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block. More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$). General-purpose compilers cannot detect this (very domain-specific) property. Our algorithm for sharing elimination will exploit this extra information.

Both CSE and factorization target spatial sharing. We will see that in the cases in which factorization is found to increase the operation count (due to the need for preliminary transformations), CCSE will instead be employed to remove spatial sharing.

%TODO : if also the Ri loop is linear, then optimality full stop.

\subsubsection{Structured inner products}
\label{sec:se-rln}
%Expressions depending on the linear loops $L_j$ and $L_k$ result from inner products involving test and trial functions (possibly, derivatives or vector-valued). The problem of reaching optimality (at least) at the level of the multilinear loop nest reduces, therefore, to understand how to optimally schedule these inner products. 
...
...
%The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients). We distinguish the following cases.
%\begin{enumerate}
%\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. This is the simplest scenario since factorizing $f_i$ has no side effects.
%\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$. Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
%\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions. This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$. We provide an example. Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
%\end{enumerate}

@@

\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization. Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression. Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.

@@

\section{Optimal Synthesis}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm capable of reaching optimality (Definition~\ref{def:optimality}) in a class of finite element integration loop nests (Definition~\ref{def:fem-loopnest}) and quasi-optimality in any other finite element integration loop nest. In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.

@@

\subsection{Trading optimality for analysis cost}
Code generation time is fundamental for productivity. In this section, we discuss how to reduce it by relaxing optimality.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints. Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials. This matter is yet to be properly investigated. However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity. The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed). For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated. An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
\end{itemize}

>>>>

\begin{Def}[Quasi-optimality of a loop nest]
\label{def:mln-quasi-optimality}
Given $\Lambda$, $\Lambda'$ and $\Gamma$ as in Definition~\ref{def:mln-optimality}, we say that $\Lambda' = \Gamma (\Lambda)$ is a quasi-optimal synthesis of $\Lambda$ if the number of operations (additions, products) that it performs to evaluate the result is at least minimal in all innermost loops.
\end{Def}

@@

Achieving optimality in polynomial time is not generally possible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured. Multilinearity, on the other hand, ensures a certain degree of regularity for the  $\alpha_{eij}$ and $\beta_{eik}$ sub-expressions. In the following, we formulate an approach that achieves: (i) optimality whenever the $\sigma_{ei}$ sub-expressions are ``sufficiently regular'';  (ii) quasi-optimality in all other cases. In particular, we build:
\begin{itemize}
\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
\item an algorithm to select points in the transformation space (Sectionm~\ref{sec:optimal-synthesis})
\end{itemize}

@@

In this section, we study \textit{sharing elimination}, a transformation that aims to reduce the operation count by removing sharing through the application of factorization, code motion and (if necessary) common sub-expression elimination. If the objective were reaching optimality and expressions were systematically unstructured, transformations of this sort would require solving large combinatorial problems -- for instance to evaluate the impact of all possible factorization strategies. Our sharing elimination strategy, however, attempts to exploit the structure that often characterizes finite element integration expressions to achieve quasi-optimality (after coordination with other transformations, as discussed in the following sections); optimality is achieved if stronger properties hold.

In Sections~\ref{sec:se-mln} and~\ref{sec:se-rln} we discuss structural (i.e., inherent in finite element integration) and potential (i.e., due to the particular use of differential and vector operators) properties of our expressions, while in Section~\ref{sec:sec:se-algo} we present the sharing elimination algorithm. We recall that we are assuming a generic finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$ with expressions in canonical form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$\footnote{A similar reasoning applies to forms and loops that are not bilinear}.

\subsubsection{Common sub-expressions}
\label{sec:se-mln}
A typical optimization to reduce the arithmetic complexity of a code is common sub-expression elimination (CSE). This often revolves around searching, possibly across different statements, and replacing with temporaries, sub-expressions that (i) are symbolically identical and (ii) reduce to the same value.  We observe that ``classic CSE'' (CCSE) can straightforwardly be applied without the need for data dependence analysis, since the body of a finite element integration loop nest is an independent basic block. More interestingly, we observe that in the ``special case'' (SCSE) in which test and trial functions are in the same function space, there may exist $w_1$ and $w_2$ such that $\alpha_{eij}^{w1} = \beta_{eik}^{w2}$, in which case common sub-expressions will arise along different loops ($L_j$ and $L_k$). General-purpose compilers cannot detect this (very domain-specific) property. Our algorithm for sharing elimination will exploit this extra information. Although both CSE and factorization target spatial sharing, the latter may actually increase the operation count if preliminary transformations (e.g., expansion) are required. In such a case, CCSE should be preferred.

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as sums of nested operations between tensors. A tensor (i.e., its entries) always depends on a specific subset of loops, and the set of loops is, obviously, always well-defined. A recurring scenario when multiplying two vectors together is that the optimal operation scheduling is to be found in a space of size 2.

Consider a generic matrix-vector product $y_{\Psi \Delta} = A_{\Psi} x_{\Delta}$, in which the entries of $y, A$ and $x$ depend on $\Psi, \Delta \subset \Lambda$. Let us assume that $y_{\Psi \Delta}$ itself is just an operand of a larger scalar-valued expression, so its entries can represent any possible sub-expression in $\lbrace \alpha_{eij}, \beta_{eik}, \sigma_{ei} \rbrace$. As an example, assume that $A$ is the inverse Jacobian matrix of a given coefficient and $x$ the gradient of a function in two dimensions. An entry in $y_{\Psi \Delta}$, therefore, is of the form $(a_{\Psi} f_{\Delta} + b_{\Psi} g_{\Delta}) (c_{\Psi} f_{\Delta} + d_{\Psi} g_{\Delta})$. To illustrate the concept, consider the case $\Psi = \emptyset$, $\Delta = \lbrace L_i \rbrace$, in which we have  $(a f_i + b g_i) (c f_i + d g_i) \in \sigma_{ei}$. To schedule this expression within $L_i$ we basically have two choices:

Many analogous examples can easily be found.

the expression $\dot(J^{-T} \nabla v, J^{-T} \nabla u)$ resulting from a Laplacian operator.

the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the coordinates field Jacobian matrix with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics

One can find this pattern in a wide range of variational forms. In fact, all of the forms used for experimentation (Section~\ref{sec:perf-results}) -- and many others that we have tried beyond these -- are characterized by this property.

In the base case, clearly, there is only one inner product between test and trial functions. In general, there may be multiple inner products, whose vectors  The problem of reaching optimality reduces to understand how to schedule these inner products in the most effective way

multilinearity -> avoids ambiguity
%TODO : if also the Ri loop is linear, then optimality full stop.
we can relax linearity by talking of structured inner products which avoids ambiguity in collection

%Expressions depending on the linear loops $L_j$ and $L_k$ result from inner products involving test and trial functions (possibly, derivatives or vector-valued). The problem of reaching optimality (at least) at the level of the multilinear loop nest reduces, therefore, to understand how to optimally schedule these inner products. 
...
...
%The arithmetic complexity of $\sigma_{ei}$ varies with the differential and vector operators employed in a monomial, becoming critical in complicated formulations (characterized, for instance, by several, possibly non-linear coefficients). We distinguish the following cases.
%\begin{enumerate}
%\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the canonical form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus \lbrace f_i \rbrace$. This is the simplest scenario since factorizing $f_i$ has no side effects.
%\item There is no common symbol among all terms, although there still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$. Factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost (i.e., code motion opportunities are destroyed).
%\item Within a term $\sigma_{ei}^{w}$, sharing elimination could still be performed by (i) factorizing integration-dependent symbols and (ii) hoisting geometry-dependent sub-expressions. This is tempting, but not necessarily the optimal choice depending on two factors: the trip count of $L_i$ and the arithmetic complexity of $\sigma_{ei}^{w}$. We provide an example. Consider the sub-expression $(a f_i + b g_i) (c f_i + d g_i)$ in $\sigma_{ei}^{w}$\footnote{The example was excerpted from the multiplication of the Jacobian matrix for the change of coordinates with the derivative of a coefficient}. We note that the form $f_i (a + c) + g_i (b + d)$ (followed by hoisting) reduces the operation count if and only if the trip count of $L_i$ is strictly greater than 1. We see that product expansion, which is essential to expose the whole space of factorizations, increases the operation count (we will see that this concept reiterates throughout the article), so a mechanism to assess the profitability of sharing elimination is required. As we detail in Algorithm~\ref{algo:opt}, the solution will consist of using combinatorics.
%\end{enumerate}

@@

\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
\label{sec:completeness}
We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization. Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression. Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.

@@

\section{Implementation Synthesis}
\label{sec:optimal-synthesis}
In this section, we build a transformation algorithm capable of reaching optimality (Definition~\ref{def:optimality}) in a class of finite element integration loop nests (Definition~\ref{def:fem-loopnest}) and quasi-optimality in any other finite element integration loop nest. In the perspective of integrating the transformation algorithm with a code generation system, we also discuss an approach that trades optimality for compilation time.

@@

\subsection{Trading operation count for analysis cost}
We can relax the constraints on (quasi-)optimality to relieve the analysis phase. In this section, we discuss ways of achieving this.
\label{sec:obsheu}
\begin{itemize}
\item \textit{Transformation algorithm.} The transformation algorithm (Figure~\ref{code:intuition}) evaluates all possible combinations in which pre-evaluation candidates can be scheduled to determine the optimal configuration honoring memory constraints. Despite the fact that we have never observed any tangible increase in analysis cost, one could expect slow-downs in complex forms with many monomials. This matter is yet to be properly investigated. However, a possibility is to avoid or constrain the search; in the worst case scenario, this would be equal to reaching local optimums.
\item \textit{Construction of $\theta$.} Instead of determining exact values for $\theta_{se}$ and $\rho$ (for $\theta_{pre}$) by (partly) executing Algorithms~\ref{algo:opt} and~\ref{algo:pre-evaluation}, approximations can be found by exploiting loop linearity. The number of unique terms in a multilinear loop nest can be used to calculate a lower bound on the innermost loop operation count (where, for higher order methods, the bulk of the computation is performed). For example, in Figure~\ref{code:multi_loopnest_a}, only one term depends on $L_j$ ($b_j$), so at most one multiplication is expected in the innermost loop once sharing is eliminated. An approximation for $\theta_{se}$ is then $\theta_{se} = n_{iters} \cdot n_{unique}$, where $n_{iters}$ is the total number of loop iterations, and $n_{unique}$ is the smallest number of unique terms depending on one the linear loops.
\end{itemize}
